---
title: "RPPA analysis development"
author: Lucy Liu
output:
  pdf_document: 
    toc: TRUE
    toc_depth: 4
  html_document: default
bibliography: rppa.bib
editor_options: 
  chunk_output_type: inline
---

\newpage

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=62, comment = FALSE), tidy=TRUE)
#chunk option to specify width of writing in chunk so it does not run off page
```

```{r, echo=FALSE}
library(RColorBrewer)

colours <-  brewer.pal(8, "Dark2")
colours[9] <- "#662506"
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(reshape2)
library(openxlsx)
library(limma)

options(digits = 3)
#global option to print 3 decimal places
```

# Introduction
The aim of this analysis is to guide the development of RPPA data analysis methodology.

Two data sets will be used to aid the analysis:

1. Robert: This was an in-house experiment comparing protein expression across various siRNA transfected conditions and two drug compound conditions in the breast cancer cell line HS578T. There are two biological replicates.
2. Stuart Gallagher: This experiment investigated various melanoma cell lines treated with DMSO and IBET151. There are two biological replicates.

# Normalisation
There are three primary factors that need to be normalised for in an RPPA experiment [@liu_comprehensive_2014,@wachter_analysis_2015]: 

1. Spatial bias: Differences in intensity caused by location of the lysate spot on the slide (e.g. rim effects). The Zepto system already accounts for this using the BSA control spots, thus spatial normalisation will not be further considered.
2. Total amount of protein (loading) of different samples on the slides: Although the total proteins in the lysate are gauged before they are printed, this is confounded by lipids and other biological materials in the samples. Thus, the total protein measurement is only a rough estimate.
3. Non linearity of variances: A MA (differences versus means) plot of the differences (M) in intensity of ABs between two samples varies across the spectrum of A, the mean intensity of the two samples. This is especially the case at the upper range and sometimes the lower range of A. 
This can be seen best when comparing a pair of technical replicates. As there would not be any differential expression, you would expect their intensities to be equal. Some imbalance often occurs and this imbalance may vary depending on the average intensity (A). Variation in the imbalance 
When normalised well all the points of a MA plot should align with a horizontal line and be evenly distributed about this line. 

## Normalisation methods
There are a number of software packages specifically developed for analysing RPPA data. The normalisation methodologies offered in each package are detail below and offer insight into the common normalisation techniques used.

* RPPanalyzer: 
    + Total protein dye
    + Housekeeping protein
    + Median normalisation
* MIRACLE
    + Housekeeping protein
    + Median normalisation
    + Variable slope

### Total protein dye
Include a total protein dye with each experiment to determine the amount of total protein in each sample. A correction factor that reflects the deviation of the protein concentration from the median concentration of all spots is calculated. AB intensities are normalised by dividing by this correction factor. This technique requires running an extra array.

### Housekeeping protein
This normalisation methodology is based on the assumption that the levels of housekeeping proteins such as $\beta$-Actin is uniform across samples and experiment conditions. Thus any differences in level of these housekeeping proteins is due to differing amounts of protein loading. To normalise to housekeeping protein(s) raw intensity values are divided by housekeeping protein RFI values.

### Loading control
This approach is utilised by MD Anderson. 

Protein effects on intensity are accounted for by dividing all the raw linear intensity by the median for each AB (across all samples). This is the 'median centered ratio'. Sample effects are accounted for by taking the median of the median centered ratios for each sample (across all ABs). This becomes the correction factor for that sample. Raw intensity values are divided by this correction factor to obtain the normalised intensity.

The steps are outlined below:

1. Determine median RFI for each AB (across all samples)
2. Divide each RFI by the median within each AB to get the 'median-centred ratio'.
3. Calculate the median median-cetered ratio for each sample (across all ABs). This is the correction factor for each sample.
4. Divide each median-centred ratio by the correction factor for each sample.

### Median normalisation/Global median centering
This method assumes that all measured proteins reflect the total protein amount of one sample. Thus the median AB of a sample estimates sample loading. The median value of all AB signals for a sample is used to normalise the raw intensity values for each AB. This is one of the 'simpliest' methods (involving the least steps) however is biased when the number of ABs is <100 [@liu_comprehensive_2014]. 

### Invariable protein normalisation
This methodology was proposed by Liu et al. [@liu_comprehensive_2014].
It proceeds thus:

1. Rank the intensity of ABs for each sample so you have ranked ABs for each sample from the highest expressing AB to lowest expressing AB.
2. Calculate the variance of the ranks for each AB. Remove the AB with the highest rank variance.
3. Re-rank the intensity of ABs.
4. Repeat the steps 2 and 3 until the number of remaining markers reaches a predetermined number (100 kept in Liu et al. study). 
5. Trim intensities for each AB e.g. the highest and lowest 25% of values are removed from the data set.
6. Average the remaining values of every AB across all samples and use this as a virtual reference sample.
7. Normalise each sample to the virtual reference sample by lowess smoothing using a MA plot. The normalised values are generated using the residuals of the fit.

The aim of this methodology is to determine the most uniformly expressed proteins and use these as an effective 'housekeeping' protein to normalise to.

### Variable slope
This methodology was developed by MD Anderson.
It aims to normalise for between array, between sample and protein effects, which result from variation in the estiamted slope parameters from the calibration curve estiamted in the quantification step. It was designed specifically for quantification (from each dilution series) via the joint sample model, 'SuperCurve'. This methodology may not translate to protein expression data not quantified using SuperCurve and will thus not be considered here.

## How to evaluate normalisation methodologies
To be able to evaluate each normalisation methodology, one must be able to compare how effectively each method normalises for total protein loading and non-linearity of variances. Normalisation of non-linearity of variances is effectively assessed using the MA plot. Normalisation of total protein loading is more difficult to assess and must be done indirectly. Two techniques have beens suggested and are discussed below.

### Pairwise correlation coefficients (Spearman $\rho$)
This plot is used to evaluate the between protein correlations. For each pair of proteins (for the 29 proteins in this experiment, there are 406 different possible pairs) the Spearman $\rho$ is calculated across all samples. This is then sorted and plotted. We expect the positive and negative correlation coefficients to be largely equal. If there is an abnormally large number of protein pairs with positive correlation, it may be due to a sample loading effect where some proteins are high in all samples, thus are ranked simiarly highly for those samples. Samples with low loading result in low levels of all proteins, and are ranked similarly low for those samples. There would thus be high correlation between a larger proportion of proteins. We expect protein expression to be inherently different due to expression, phosphorylation levels and/or AB affinity. 

### Distrubution of sample medians
To evaluation effectiveness of loading correction, the distribution of the median raw intensity values for each sample should have a narrow dispersion - i.e. if protein loading is normalised for, you would not expect the median raw intensity value of your samples to vary significantly.

## Robert Vary data set
The Robert data set will be normalised using various methods. The performance of each method will be evaulated.

The details of sample conditions are detailed below:
```{r}
sampleRob <- read.xlsx("Data/RunSummary_SIMPSON_RoberyVary_R067_20180521.xlsx",
                         sheet = 4, colNames = TRUE, rowNames = FALSE)
kable(sampleRob)
```

Read in the raw normalised to secondary AB RFI values and print the beginning of the data:
```{r}
raw <- read.xlsx("Data/RunSummary_SIMPSON_RoberyVary_R067_20180521.xlsx", 
                 sheet = 2, colNames = TRUE, rowNames = TRUE)
#read in raw data
raw <- raw[-1,]
#remove the second row, which contained in-house AB names
raw[] <- sapply(raw, FUN = as.double)
#convert all rows into double
raw <- t(raw)
#transpose (swap rows and columns)
raw[1:5,1:9]
```

### Raw intensity values
Visualise raw intensity values by AB via bloxplot:
```{r}
raw_tidy <- melt(raw)
colnames(raw_tidy) <- c("AB", "Sample", "RFI")
#Convert raw matrix into tidy data format
```
```{r}
ggplot(raw_tidy, aes(y = RFI, x = AB)) + 
  geom_boxplot() + 
  coord_flip()
```

Remove "CD44" and "Vimentin" and remake boxplots to better visualise the other AB's.
```{r}
raw_tidy %>%
  filter(! AB %in% c("CD44", "Vimentin")) %>%
  ggplot(aes(y = RFI, x = AB)) + 
  geom_boxplot() + 
  coord_flip()
```

Visualise raw intensity values per sample via boxplot:
```{r}
ggplot(raw_tidy, aes(y = RFI, x = Sample)) + 
  geom_boxplot() + 
  coord_flip()
```

### Housekeeping normalisation

The housekeeping proteins in this experiment were $\beta$-Actin and Prohibitin. 

Normalisation against $\beta$-Actin, Prohibitin and the mean of these two housekeeping proteins will all be performed. The results will then be compared.

```{r}
hk <- raw_tidy %>%
  group_by(Sample) %>%
  mutate(HK_B = RFI / RFI[AB == "Beta.Actin"]) %>%
  mutate(HK_P = RFI / RFI[AB == "Prohibitin"]) %>%
  mutate(meanHK = mean(RFI[AB %in% c("Beta.Actin", "Prohibitin")])) %>%
  ungroup() %>%
  mutate(HK_mean = RFI / meanHK) %>%
  select(-meanHK)

hk <- hk %>%
  gather(HK_B, HK_P, HK_mean, RFI, key = "Type", value = "Intensity")
```

Plot the distribution of the intensity values for each of the normalisation methods and the raw RFI. 

```{r}
ggplot(hk, aes(x = Intensity, colour = Type)) +
  geom_density()
```

Zoom in on the low intensity values to better visualise the data:
```{r}
ggplot(hk, aes(x = Intensity, colour = Type)) +
  geom_density() +
  coord_cartesian(xlim = c(0,5))
```

The distribution of the normalised intensity values are very similar for all three housekeeping normalisation methods. Overall, housekeeping normalisation has reduced the number of ABs with low intensity values.

Visualise the change in intensity distribution for each sample:
```{r, out.width="33%"}
samples <- unique(hk$Sample)
for (i in 1:26){
print(
  hk %>%
  filter(Sample == samples[i]) %>%
ggplot(aes(x = Intensity, colour = Type, y = ..scaled..)) +
  geom_density() 
)
}
```

Plot the median raw intensity values per sample:
```{r}
hk %>%
  group_by(Sample, Type) %>%
  summarise(sampleMed = median(Intensity, na.rm = TRUE)) %>%
  ggplot(aes(x = sampleMed, colour = Type)) + 
  geom_density()
```

The sample median intensity values are higher (as suggested by the distribution of intensity values above) and more spread out for the housekeeping normalised values when compared to the raw intensity values.

MA plots of the difference and average intensity levels of each pair of biological replicates are plotted below. 
Each point represents a protein. You expect the points to lie along 0 and the variation of points around 0 to be similar along A.

As all three housekeeping normalisations are similar, we will only compare raw values with that normalised to the mean of $\beta$-Actin and Prohibitin.
```{r, out.width="50%"}
raw_matrix <- log(as.matrix(raw) + 1e-5, base = 2)
hk_matrix <- acast(
  hk %>%
  filter(Type == "HK_mean") %>%
  mutate(log2 = log(Intensity + 1e-5, base = 2)) %>%
  select(-c(Type, Intensity)),
  
  AB~Sample,
  value.var = "log2"
)

sampleNames <- colnames(hk_matrix)

for (i in 1:13){
plotMA(raw_matrix[,(2*i-1):(2*i)], main = paste(sampleNames[i], "Raw", sep="-")) 
plotMA(hk_matrix[,(2*i-1):(2*i)], main = paste(sampleNames[i], "Housekeeping normalised", sep="-"))
}
```

A spearman rank correlation plot allows us to evaluate the between protein correlations. We expect a similar number of proteins with > 0.5 and < -0.5 correlations.  
Blue points are the correlations between proteins for raw RFI values. Black are correlations between proteins for the values normalised to mean of the two housekeeping proteins.
```{r, echo=FALSE}
spearmanRanks <- function(rankdf, c1){

  # Takes in a df where there is a column called "AB" (which contains the names of ABs) and another column of ranks of each AB across all samples and returns a vector of spearman correlations between the ranks of all possible combinations of ABs.
  
  
  # rankdf = dataframe of AB ranks. Each AB is ranked according to expression,     across all samples.
  # c1 = Name of the column containg the ranks. 


  ABcomb <- combn(unique(rankdf$AB),2)
  # Create matrix of all possible combinations. Resulting matrix will have 2 rows   as we are choosing 2 from vector of unique ABs.
  
  c <- vector(mode = "double", length = dim(ABcomb)[2])
  # Create empty vector length of the number of combinations
  
  for (i in 1:dim(ABcomb)[2]){
  # Cycle i through the number of combinations
    
  a <- rankdf[[c1]][rankdf$AB == ABcomb[1,i]]
  # Note cannot use rankdf$c1 as it will look for the column named "c1". Used instead rankdf[[c1]]. Double brackets returns vector instead of df/tibble
  b <- rankdf[[c1]][rankdf$AB == ABcomb[2,i]]
  
  c[i] <- cor(a,b, method = 'spearman')
  }
  
  return(c)
  
}

spearmanPlot <- function(rankdf, c1, legend){
  
  a <- spearmanRanks(rankdf = rankdf, c1 = as.character(c1))
  # gives you vector of correlations for the normalisation method of interest
  b <- spearmanRanks(rankdf = rankdf, c1 = "RFI")
  # gives you vector of correlations for the raw RFI values
  plot(a[order(a)], ylim = c(-1,1))
  points(b[order(b)], col = "Blue")
  legend("bottomright", legend = c(legend, "Raw"), text.col = c("Black", "Blue"))
  abline(0.5,0)
  abline(-0.5,0)
  # plot the ordered values.
  
}
```
```{r}
rank_hk <- hk %>%
  group_by(Type,AB) %>%
  mutate(Rank = rank(Intensity)) %>%
  select(-Intensity) %>%
  spread(key = Type, value = Rank)

spearmanPlot(rank_hk, "HK_mean", "Housekeeping")
```

After normalisation, there is a much more even spread of correlations above and below 0, as we would expect for data well normalised for total protein loading. 

### Loading control

Calculate normalised values using loading control method:
```{r}
lc <- raw_tidy %>%
  group_by(AB) %>%
  mutate(medianAB = median(RFI)) %>%
  #Find median RFI for each AB
  ungroup() %>%
  mutate(medCenter = RFI/medianAB) %>%
  #Divide RFI by the median AB, for each AB
  group_by(Sample) %>%
  mutate(CF = median(medCenter, na.rm = TRUE)) %>%
  #Find the median medCenter for each sample. Remove NA values
  ungroup() %>%
  mutate(LC = medCenter/CF) %>%
  select(c(AB, Sample, RFI, LC))

lc <- lc %>%
  gather(LC, RFI, key = "Type", value = "Intensity")
```

Plot distribution of all intensity values for raw and loading control normalised data:
```{r, message=FALSE}
ggplot(lc, aes(x = Intensity, colour = Type)) +
  geom_density()
```

Zoom in on the low intensity values to better visualise the data:
```{r}
ggplot(lc, aes(x = Intensity, colour = Type)) +
  geom_density() +
  coord_cartesian(xlim = c(0,3))
```

```{r, out.width="33%", warning=FALSE}
samples <- unique(lc$Sample)
for (i in 1:26){
print(
  lc %>%
  filter(Sample == samples[i]) %>%
ggplot(aes(x = Intensity, colour = Type, y = ..scaled..)) +
  geom_density() + 
  labs(title = samples[i])
)
}
```

LC normalisation does not appear to significantly change the overall distribution of intensity values.

LC normalisation results in the median sample intensities to all be 1 thus the distribution of sample medians is not useful in evaluating its performance.

Plot MA graphs for raw and loading control normalised intensities:
```{r, out.width="50%"}
raw_matrix <- log(as.matrix(raw) + 1e-5, base = 2)

lc_matrix <- acast(
  lc %>%
  filter(Type == "LC") %>%
  mutate(log2 = log(Intensity + 1e-5, base = 2)) %>%
  select(-c(Type, Intensity)),
  
  AB~Sample,
  value.var = "log2"
)

sampleNames <- colnames(lc_matrix)

for (i in 1:13){
plotMA(raw_matrix[,(2*i-1):(2*i)], main = paste(sampleNames[i], "Raw", sep="-")) 
plotMA(lc_matrix[,(2*i-1):(2*i)], main = paste(sampleNames[i], "Loading control", sep="-"))
}
```

Spearman rank correlation plot. We expect a largely equal psotive and negative correlation as discussed above.
Blue points are the correlations between proteins for raw RFI values. Black are correlations between proteins for loading control normalised values.
```{r}
rank_lc <- lc %>%
  group_by(Type,AB) %>%
  mutate(Rank = rank(Intensity)) %>%
  select(-Intensity) %>%
  spread(key = Type, value = Rank)

spearmanPlot(rank_lc, "LC", "Loading Control")
```

Again, there is a more even distribution of correlation above and below 0 after loading control normalisation.

### Median normalisation/Global median centering

Calculate median normalised values.
```{r}
mn <- raw_tidy %>%
  group_by(Sample) %>%
  mutate(med = median(RFI)) %>%
  ungroup() %>%
  mutate(MN = RFI/med) %>%
  select(-med)

mn <- mn %>%
  gather(RFI, MN, key = "Type", value = "Intensity")
```

Plot distribution of all intensity values for raw and median normalised data:
```{r}
ggplot(mn, aes(x = Intensity, colour = Type)) +
  geom_density()
```

Zoom in on the low intensity values to better visualise the data:
```{r}
ggplot(mn, aes(x = Intensity, colour = Type)) +
  geom_density() +
  coord_cartesian(xlim = c(0,5))
```

Median normalisation increases intensity values, even more significantly than HK normalisation.

Plot the distribution of intensity values for each sample:
```{r, out.width="33%"}
samples <- unique(mn$Sample)
for (i in 1:26){
print(
  mn %>%
  filter(Sample == samples[i]) %>%
ggplot(aes(x = Intensity, colour = Type, y = ..scaled..)) +
  geom_density() 
)
}
```

Distribution of sample medians for raw and median normalised intensities:
```{r}
mn %>%
  group_by(Sample, Type) %>%
  summarise(sampleMed = median(Intensity, na.rm = TRUE)) %>%
  ggplot(aes(x = sampleMed, colour = Type)) + 
  geom_density()
```

Plot MA graphs for raw and median normalised intensities:
```{r, out.width="50%"}
mn_matrix <- acast(
  mn %>%
  filter(Type == "MN") %>%
  mutate(log2 = log(Intensity + 1e-5, base = 2)) %>%
  select(-c(Type, Intensity)),
  
  AB~Sample,
  value.var = "log2"
)

sampleNames <- colnames(mn_matrix)

for (i in 1:13){
plotMA(raw_matrix[,(2*i-1):(2*i)], main = paste(sampleNames[i], "Raw", sep="-")) 
plotMA(mn_matrix[,(2*i-1):(2*i)], main = paste(sampleNames[i], "Median normalised", sep="-"))
}
```

Spearman rank correlation plot. We expect a largely equal postive and negative correlation as discussed above.
Blue points are the correlations between proteins for raw RFI values. Black are correlations between proteins for median normalised values.
```{r}
rank_mn <- mn %>%
  group_by(Type,AB) %>%
  mutate(Rank = rank(Intensity)) %>%
  select(-Intensity) %>%
  spread(key = Type, value = Rank)

spearmanPlot(rank_mn, "MN", "Median normalised")
```

There is a much larger proportion of protein pairs with correlation >0.5 for both raw and median normalised data. 

### MDS plots
Distances on an MDS plot correspond to leading log-fold change between each pair of samples. MDS plots for raw data and each normalisation methodology is plotted below.

#### Raw

```{r}
groupCond <- as.factor(sampleRob$Condition)
group_col <- groupCond
levels(group_col) <- colours[1:9]
plotMDS(raw_matrix, col = as.character(group_col), main = "MDS coloured by condition", xlim = c(-1.5,2))
legend("topright", 
       legend = levels(groupCond), 
       text.col = levels(group_col), bty = "n")
```

```{r}
groupTime <- as.factor(sampleRob$TimePoint)
group_col <- groupTime
levels(group_col) <- colours[1:2]
plotMDS(raw_matrix, col = as.character(group_col), main = "MDS coloured by Time point")
legend("topright", 
       legend = levels(groupTime), 
       text.col = levels(group_col), bty = "n")
```

#### Housekeeping

```{r, echo=FALSE}
colnames(hk_matrix) == sampleRob$Sample
colnames(lc_matrix) == sampleRob$Sample
colnames(mn_matrix) == sampleRob$Sample
colnames(raw_matrix) == sampleRob$Sample
#Check that the sample names in the data matrix match that of the phenotype dataframe.
```

```{r}
groupCond <- as.factor(sampleRob$Condition)
group_col <- groupCond
levels(group_col) <- colours[1:9]
plotMDS(hk_matrix, col = as.character(group_col), main = "MDS coloured by condition", xlim = c(-1.8,2.5), ylim = c(-1.5,1.5))
legend("topright", 
       legend = levels(groupCond), 
       text.col = levels(group_col), bty = "n")
```

```{r}
groupTime <- as.factor(sampleRob$TimePoint)
group_col <- groupTime
levels(group_col) <- colours[1:2]
plotMDS(hk_matrix, col = as.character(group_col), main = "MDS coloured by Time point", xlim = c(-1.8,2.5), ylim = c(-1.5,1.8))
legend("topright", 
       legend = levels(groupTime), 
       text.col = levels(group_col), bty = "n")
```

#### Loading control

```{r}
groupCond <- as.factor(sampleRob$Condition)
group_col <- groupReps
levels(group_col) <- colours[1:9]
plotMDS(lc_matrix, col = as.character(group_col), main = "MDS coloured by condition", xlim = c(-4,4.5))
legend("topright", 
       legend = levels(groupReps), 
       text.col = levels(group_col), bty = "n")
```

```{r}
groupTime <- as.factor(sampleRob$TimePoint)
group_col <- groupTime
levels(group_col) <- colours[1:2]
plotMDS(lc_matrix, col = as.character(group_col), main = "MDS coloured by Time point", xlim = c(-4,3))
legend("topright", 
       legend = levels(groupTime), 
       text.col = levels(group_col), bty = "n")
```

#### Median normalisation

```{r}
groupCond <- as.factor(sampleRob$Condition)
group_col <- groupReps
levels(group_col) <- colours[1:9]
plotMDS(mn_matrix, col = as.character(group_col), main = "MDS coloured by condition", xlim = c(-2.2,2.5))
legend("topright", 
       legend = levels(groupReps), 
       text.col = levels(group_col), bty = "n")
```

```{r}
groupTime <- as.factor(sampleRob$TimePoint)
group_col <- groupTime
levels(group_col) <- colours[1:2]
plotMDS(mn_matrix, col = as.character(group_col), main = "MDS coloured by Time point", xlim = c(-2.2,2.5))
legend("topright", 
       legend = levels(groupTime), 
       text.col = levels(group_col), bty = "n")
```

## Stuart Gallagher
The Stuart Gallagher data set we will normalise according to various methods and evaulate their performance.

melanoma cell 52. ibet

The details of sample conditions are detailed below:
```{r}
sample_info <- read.xlsx("Data/RunSummary_SIMPSON_RoberyVary_R067_20180521.xlsx",
                         sheet = 4, colNames = TRUE, rowNames = FALSE)
kable(sample_info)
```

Read in the raw normalised to secondary AB RFI values and print the beginning of the data:
```{r}
raw <- read.xlsx("Data/RunSummary_SIMPSON_RoberyVary_R067_20180521.xlsx", 
                 sheet = 2, colNames = TRUE, rowNames = TRUE)
#read in raw data
raw <- raw[-1,]
#remove the second row, which contained in-house AB names
raw[] <- sapply(raw, FUN = as.double)
#convert all rows into double
raw <- t(raw)
#transpose (swap rows and columns)
raw[1:5,1:9]
```

### Raw intensity values
Visualise raw intensity values by AB via bloxplot:
```{r}
raw_tidy <- melt(raw)
colnames(raw_tidy) <- c("AB", "Sample", "RFI")
#Convert raw matrix into tidy data format
```
```{r}
ggplot(raw_tidy, aes(y = RFI, x = AB)) + 
  geom_boxplot() + 
  coord_flip()
```

Remove "CD44" and "Vimentin" and remake boxplots to better visualise the other AB's.
```{r}
raw_tidy %>%
  filter(! AB %in% c("CD44", "Vimentin")) %>%
  ggplot(aes(y = RFI, x = AB)) + 
  geom_boxplot() + 
  coord_flip()
```

Visualise raw intensity values per sample via boxplot:
```{r}
ggplot(raw_tidy, aes(y = RFI, x = Sample)) + 
  geom_boxplot() + 
  coord_flip()
```

### Housekeeping normalisation
The housekeeping proteins in this experiment were $\beta$-Actin and Prohibitin. 

Normalisation against $\beta$-Actin, Prohibitin and the mean of these two housekeeping proteins will all be performed. The results will then be compared.

```{r}
hk <- raw_tidy %>%
  group_by(Sample) %>%
  mutate(HK_B = RFI / RFI[AB == "Beta.Actin"]) %>%
  mutate(HK_P = RFI / RFI[AB == "Prohibitin"]) %>%
  mutate(meanHK = mean(RFI[AB %in% c("Beta.Actin", "Prohibitin")])) %>%
  ungroup() %>%
  mutate(HK_mean = RFI / meanHK) %>%
  select(-meanHK)

hk <- hk %>%
  gather(HK_B, HK_P, HK_mean, RFI, key = "Type", value = "Intensity")
```

Plot the distribution of the intensity values for each of the normalisation methods and the raw RFI. 

```{r}
ggplot(hk, aes(x = Intensity, colour = Type)) +
  geom_density()
```

Zoom in on the low intensity values to better visualise the data:
```{r}
ggplot(hk, aes(x = Intensity, colour = Type)) +
  geom_density() +
  coord_cartesian(xlim = c(0,5))
```

The distribution of the normalised intensity values are very similar for all three housekeeping normalisation methods. Overall, housekeeping normalisation has reduced the number of ABs with low intensity values.

Visualise the change in intensity distribution for each sample:
```{r, out.width="33%"}
samples <- unique(hk$Sample)
for (i in 1:26){
print(
  hk %>%
  filter(Sample == samples[i]) %>%
ggplot(aes(x = Intensity, colour = Type, y = ..scaled..)) +
  geom_density() 
)
}
```

Plot the median raw intensity values per sample:
```{r}
hk %>%
  group_by(Sample, Type) %>%
  summarise(sampleMed = median(Intensity, na.rm = TRUE)) %>%
  ggplot(aes(x = sampleMed, colour = Type)) + 
  geom_density()
```

The sample median intensity values are higher (as suggested by the distribution of intensity values above) and more spread out for the housekeeping normalised values when compared to the raw intensity values.

MA plots of the difference and average intensity levels of each pair of biological replicates are plotted below. 
Each point represents a protein. You expect the points to lie along 0 and the variation of points around 0 to be similar along A.

As all three housekeeping normalisations are similar, we will only compare raw values with that normalised to the mean of $\beta$-Actin and Prohibitin.
```{r, out.width="50%"}
raw_matrix <- log(as.matrix(raw) + 1e-5, base = 2)
hk_matrix <- acast(
  hk %>%
  filter(Type == "HK_mean") %>%
  mutate(log2 = log(Intensity + 1e-5, base = 2)) %>%
  select(-c(Type, Intensity)),
  
  AB~Sample,
  value.var = "log2"
)

for (i in 1:13){
plotMA(raw_matrix[,(2*i-1):(2*i)], main = "Raw") 
plotMA(hk_matrix[,(2*i-1):(2*i)], main = "Housekeeping normalised")
}
```

A spearman rank correlation plot allows us to evaluate the between protein correlations. We expect a similar number of proteins with >0.5 and <-0.5 correlations.  
Blue points are the correlations between proteins for raw RFI values. Black are correlations between proteins for the values normalised to mean of the two housekeeping proteins.
```{r, echo=FALSE}
spearmanRanks <- function(rankdf, c1){
  
  ABcomb <- combn(unique(rankdf$AB),2)
  #Create matrix of all possible combinations. Resulting matrix will have 2 rows as   #we are choosing 2 from vector of unique ABs.
  
  c <- vector(mode = "double", length = dim(ABcomb)[2])
  #Create empty vector length of the number of combinations
  
  for (i in 1:dim(ABcomb)[2]){
  #Cycle i through the number of combinations
    
  a <- rankdf[[c1]][rankdf$AB == ABcomb[1,i]]
  #Note cannot use rankdf$c1 as it will look for the column named "c1". Used instead   #rankdf[[c1]]. Double brackets returns vector instead of df/tibble
  b <- rankdf[[c1]][rankdf$AB == ABcomb[2,i]]
  
  c[i] <- cor(a,b, method = 'spearman')
  }
  
  return(c)
  
}

spearmanPlot <- function(rankdf, c1, legend){
  
  a <- spearmanRanks(rankdf = rankdf, c1 = as.character(c1))
  b <- spearmanRanks(rankdf = rankdf, c1 = "RFI")
  plot(a[order(a)], ylim = c(-1,1))
  points(b[order(b)], col = "Blue")
  legend("bottomright", legend = c(legend, "Raw"), text.col = c("Black", "Blue"))
  abline(0.5,0)
  abline(-0.5,0)
  
}
```
```{r}
rank_hk <- hk %>%
  group_by(Type,AB) %>%
  mutate(Rank = rank(Intensity)) %>%
  select(-Intensity) %>%
  spread(key = Type, value = Rank)


spearmanPlot(rank_hk, "HK_mean", "Housekeeping")
```

After normalisation, there is a much more even spread of correlations above 0.5 and below -0.5, as we would expect for data well normalised for total protein loading. 

### Loading control
Calculate normalised values using loading control method:
```{r}
lc <- raw_tidy %>%
  group_by(AB) %>%
  mutate(medianAB = median(RFI)) %>%
  #Find median RFI for each AB
  ungroup() %>%
  mutate(medCenter = RFI/medianAB) %>%
  #Divide RFI by the median AB, for each AB
  group_by(Sample) %>%
  mutate(CF = median(medCenter, na.rm = TRUE)) %>%
  #Find the median medCenter for each sample. Remove NA values
  ungroup() %>%
  mutate(LC = medCenter/CF) %>%
  select(c(AB, Sample, RFI, LC))

lc <- lc %>%
  gather(LC, RFI, key = "Type", value = "Intensity")
```

Plot distribution of all intensity values for raw and loading control normalised data:
```{r, message=FALSE}
ggplot(lc, aes(x = Intensity, colour = Type)) +
  geom_density()
```

Zoom in on the low intensity values to better visualise the data:
```{r}
ggplot(lc, aes(x = Intensity, colour = Type)) +
  geom_density() +
  coord_cartesian(xlim = c(0,3))
```

```{r, out.width="33%", warning=FALSE}
samples <- unique(lc$Sample)
for (i in 1:26){
print(
  lc %>%
  filter(Sample == samples[i]) %>%
ggplot(aes(x = Intensity, colour = Type, y = ..scaled..)) +
  geom_density() + 
  labs(title = samples[i])
)
}
```

LC normalisation does not appear to significantly change the overall distribution of intensity values.

LC normalisation results in the median sample intensities to all be 1 thus the distribution of sample medians is not useful in evaluating its performance.

Plot MA graphs for raw and loading control normalised intensities:
```{r, out.width="50%"}
raw_matrix <- log(as.matrix(raw) + 1e-5, base = 2)

lc_matrix <- acast(
  lc %>%
  filter(Type == "LC") %>%
  mutate(log2 = log(Intensity + 1e-5, base = 2)) %>%
  select(-c(Type, Intensity)),
  
  AB~Sample,
  value.var = "log2"
)

for (i in 1:13){
plotMA(raw_matrix[,(2*i-1):(2*i)], main = "Raw") 
plotMA(lc_matrix[,(2*i-1):(2*i)], main = "Loading control normalised")
}
```

Spearman rank correlation plot. We expect a largely equal psotive and negative correlation as discussed above.
Blue points are the correlations between proteins for raw RFI values. Black are correlations between proteins for loading control normalised values.
```{r}
rank_lc <- lc %>%
  group_by(Type,AB) %>%
  mutate(Rank = rank(Intensity)) %>%
  select(-Intensity) %>%
  spread(key = Type, value = Rank)

spearmanPlot(rank_lc, "LC", "Loading Control")
```

Again, there is a more even distribution of correlation above 0.5 and below -0.5 after loading control normalisation.

### Median normalisation/Global median centering

Calculate median normalised values.
```{r}
mn <- raw_tidy %>%
  group_by(Sample) %>%
  mutate(med = median(RFI)) %>%
  ungroup() %>%
  mutate(MN = RFI/med) %>%
  select(-med)

mn <- mn %>%
  gather(RFI, MN, key = "Type", value = "Intensity")
```

Plot distribution of all intensity values for raw and median normalised data:
```{r}
ggplot(mn, aes(x = Intensity, colour = Type)) +
  geom_density()
```

Zoom in on the low intensity values to better visualise the data:
```{r}
ggplot(mn, aes(x = Intensity, colour = Type)) +
  geom_density() +
  coord_cartesian(xlim = c(0,5))
```

Median normalisation increases intensity values, even more significantly than HK normalisation.

Plot the distribution of intensity values for each sample:
```{r, out.width="33%"}
samples <- unique(mn$Sample)
for (i in 1:26){
print(
  mn %>%
  filter(Sample == samples[i]) %>%
ggplot(aes(x = Intensity, colour = Type, y = ..scaled..)) +
  geom_density() 
)
}
```

Distribution of sample medians for raw and median normalised intensities:
```{r}
mn %>%
  group_by(Sample, Type) %>%
  summarise(sampleMed = median(Intensity, na.rm = TRUE)) %>%
  ggplot(aes(x = sampleMed, colour = Type)) + 
  geom_density()
```

Plot MA graphs for raw and median normalised intensities:
```{r, out.width="50%"}
mn_matrix <- acast(
  mn %>%
  filter(Type == "MN") %>%
  mutate(log2 = log(Intensity + 1e-5, base = 2)) %>%
  select(-c(Type, Intensity)),
  
  AB~Sample,
  value.var = "log2"
)

for (i in 1:13){
plotMA(raw_matrix[,(2*i-1):(2*i)], main = "Raw") 
plotMA(mn_matrix[,(2*i-1):(2*i)], main = "Median normalised")
}
```

Spearman rank correlation plot. We expect a largely equal positive and negative correlation as discussed above.
Blue points are the correlations between proteins for raw RFI values. Black are correlations between proteins for median normalised values.
```{r}
rank_mn <- mn %>%
  group_by(Type,AB) %>%
  mutate(Rank = rank(Intensity)) %>%
  select(-Intensity) %>%
  spread(key = Type, value = Rank)

spearmanPlot(rank_mn, "MN", "Median normalised")
```

There is a much larger proportion of protein pairs with correlation >0.5 for both raw and median normalised data. 



# References

